Enzo+Moray: Radiation Hydrodynamics Adaptive Mesh Refinement Simulations with Adaptive Ray Tracing

### 2010-08-12 ms.tex
Minor changes to the timestepping section.  Adding .hgignore and bbl file.

There have been several methods of choosing a timestep to solve
radiation transfer equation because an accurate yet large timestep is
not trivial to compute.  With a small enough timestep, the solution is
guaranteed to avoid any inaccuracies (ignoring any systematic ones)
due to timestepping, but the solver must be called many times.  These
frequent calls may be unneccesary because the same solution may be
accomplished with a longer timestep.  Furthermore with ray tracing,
the photon packages only advance a short distance, and they will exist
in every $dx/dt_P$ cells with radiation and are stored between
timesteps, excessively consuming memory.  If the timestep is too
large, the solution will become inaccurate; specifically, ionization
fronts will advance too slowly, as radiation intensity exponentially
drops with a scale length of the mean free path
\begin{equation}
  \label{eqn:mfp}
  \lambda_{\rm mfp} = \frac{1}{n_{\rm abs} \; \sigma_{\rm abs}}
\end{equation}
past the ionization front.  For example in our implementation, the
chemo-thermal state of the system will remain constant as the rays are
traced through the cells.  Thus in the case of a single \ion{H}{2}
region, the speed of the ionization front is limited to approximately
the $\lambda_{\rm mfp} / dt_P$.  We can investigate this further by
evaluating the ionization front velocity in a growing Str\"{o}mgren
sphere, where $\dot{N}_\gamma = 4 \pi R^2 n_{\rm H} v_{\rm IF}$.
Using $\kph \propto \dot{n}_{\rm H} / n_{\rm H}$ and $\kph =
\dot{N}_\gamma \sigma / 4\pi R^2 A_{\rm cell}$, we can make
substitutions on both sides of the equation to arrive at the
ionization front velocity
%
\begin{equation}
  \label{eqn:vifront1}
  v_{\rm IF} \propto \frac{n_{\rm H}}{\dot{n}_{\rm H}}.
\end{equation}

On the shortest timescale, one can safely set the timestep to the
light-crossing time of a cell \citep{Abel99_RT, Trac07} but encounters
the problems stated above.  Next we describe several methods to
calculate the radiative transfer timestep.

@@

Another strategy is only allowing the neutral fraction of the ionized
gas to change a small amount, i.e. for a single cell,
%
\begin{equation}
  \label{eqn:dtelec}
  dt_{\rm P,cell} = \epsilon_{\rm ion} \frac{n_{\rm HI}}{|dn_{\rm HI}/dt|} =
  \frac{\epsilon_{\rm ion}}{|k_{\rm ph} + n_e (C_{\rm H} + \alpha_B)|},
\end{equation}
where $\epsilon_{\rm ion}$ is the maximum fraction change in neutral
fraction.  \citet{Shapiro04} found that this limited the speed of the
ionization front for the same reasons described in Equation
\ref{eqn:vifront1}.  We have implemented this method but we only
consider cells within the ionization front (by experiment we choose
$\tau > 0.5$) because we are interested in evolving ionization fronts
at the correct speed.  In the ionized region, the absolute changes in
neutral fraction are small and will not significantly affect the
ionization front evolution.  In other words, $n_{\rm HI}/(dn_{\rm
  HI}/dt)$ may be large but $(dn_{\rm HI}/dt) \sim 0$, thus we can
safely ignore these cells when determining the timestep without
sacrificing accuracy.

@@

To make this approach computationally straightforward, we consider a
spherically symmetric case without recombinations.  Thus all photons
emitted from the source result in an ionization (see
Eq. \ref{eqn:vifront1} and its derivation).  Ultimately in any
approach, we are focused on accurately following the time evolution of
the radiation field.  In this method, we base the timestep directly on
the change in radiation field, the specific intensity, $I/\dot{I}$.
We consider the field after propagating through the cell, so $I = I_0
\exp(-\tau)$, where $\tau = n_{\rm H} \sigma dl$ is the optical depth
through the cell.  The change in specific intensity is
%
\begin{equation}
  \label{eqn:vifront2}
    \frac{dI}{dt} = I_0 \exp(-\tau) (-\dot{n}_{\rm H} \sigma dl),
\end{equation}
%
which can be expressed in terms of local optical depth and neutral
fraction,
\begin{equation}
  \label{eqn:vifront3}
    \frac{dI}{dt} = -I \frac{\tau \dot{n}_{\rm H}}{n_{\rm H}}.
\end{equation}
%
This results in a local timestep
%
\begin{equation}
  \label{eqn:vifront4}
  dt_P = \frac{I}{|dI/dt|} = C_{\rm RT,cfl} \frac{n_{\rm H}}{\tau \dot{n}_{\rm H}},
\end{equation}
%
where $C_{\rm RT,cfl}$ is a safety factor that restricts the change in
intensity to its inverse.  In practice, we have found that a ceiling
of 3 can be placed on the optical depth, so optically thick cells do
not create a very small timestep.  We still find excellent agreement
with analytical solutions with this approximation.  We show the
accuracy using this timestep method in \S\ref{sec:dt_dep}.

>>>>

There have been several methods of choosing a timestep to solve
radiation transfer equation because an accurate yet large timestep is
not trivial to compute.  We describe several methods to calculate the
radiative transfer timestep in this section.  With a small enough
timestep, the solution is guaranteed to avoid any inaccuracies
(ignoring any systematic ones) due to timestepping, but the solver
must be called many times.  These frequent calls may be unneccesary
because the same solution may be accomplished with a longer timestep.
Furthermore with ray tracing, the photon packages only advance a short
distance, and they will exist in every $dx/dt_P$ cells with radiation
and are stored between timesteps, excessively consuming memory.  On
the shortest timescale, one can safely set the timestep to the
light-crossing time of a cell \citep{Abel99_RT, Trac07} but encounters
the problems stated above.

If the timestep is too large, the solution will become inaccurate;
specifically, ionization fronts will advance too slowly, as radiation
intensity exponentially drops with a scale length of the mean free
path
\begin{equation}
  \label{eqn:mfp}
  \lambda_{\rm mfp} = \frac{1}{n_{\rm abs} \; \sigma_{\rm abs}}
\end{equation}
past the ionization front.  For example in our implementation, the
chemo-thermal state of the system remains constant as the rays are
traced through the cells.  Thus in the case of a single \ion{H}{2}
region, the speed of the ionization front is limited to approximately
the $\lambda_{\rm mfp} / dt_P$.

@@

Another strategy is only allowing the neutral fraction of the ionized
gas to change a small amount, i.e. for a single cell,
%
\begin{equation}
  \label{eqn:dtelec}
  dt_{\rm P,cell} = \epsilon_{\rm ion} \frac{n_{\rm HI}}{|dn_{\rm HI}/dt|} =
  \frac{\epsilon_{\rm ion}}{|k_{\rm ph} + n_e (C_{\rm H} + \alpha_B)|},
\end{equation}
where $\epsilon_{\rm ion}$ is the maximum fraction change in neutral
fraction.  \citet{Shapiro04} found that this limited the speed of the
ionization front.  We can investigate this further by evaluating the
ionization front velocity in a growing Str\"{o}mgren sphere without
recombinations, where $\dot{N}_\gamma = 4 \pi R^2 n_{\rm H} v_{\rm
  IF}$.  Using $\kph \propto \dot{n}_{\rm H} / n_{\rm H}$ and $\kph =
\dot{N}_\gamma \sigma / 4\pi R^2 A_{\rm cell}$, we can make
substitutions on both sides of the equation to arrive at the
ionization front velocity $v_{\rm IF} \propto n_{\rm H} /\dot{n}_{\rm
  H}$.

We have implemented this method but we only consider cells within the
ionization front (by experiment we choose $\tau > 0.5$) because we are
interested in evolving ionization fronts at the correct speed.  In the
ionized region, the absolute changes in neutral fraction are small and
will not significantly affect the ionization front evolution.  In
other words, $n_{\rm HI}/(dn_{\rm HI}/dt)$ may be large but $(dn_{\rm
  HI}/dt) \sim 0$, thus we can safely ignore these cells when
determining the timestep without sacrificing accuracy.

@@

To make this approach computationally straightforward, we consider a
spherically symmetric case without recombinations.  Thus all photons
emitted from the source result in an ionization (also see
\S\ref{sec:dt_hi}).  Ultimately in any approach, we are focused on
accurately following the time evolution of the radiation field.  In
this method, we base the timestep directly on the change in radiation
field, the specific intensity, $I/\dot{I}$.  We consider the field
after propagating through the cell, so $I = I_0 \exp(-\tau)$, where
$\tau = n_{\rm H} \sigma dl$ is the optical depth through the cell.
The change in specific intensity is
%
\begin{equation}
  \label{eqn:vifront2}
    \frac{dI}{dt} = I_0 \exp(-\tau) (-\dot{n}_{\rm H} \sigma dl),
\end{equation}
%
which can be expressed in terms of local optical depth and neutral
fraction,
\begin{equation}
  \label{eqn:vifront3}
    \frac{dI}{dt} = -I \frac{\tau \dot{n}_{\rm H}}{n_{\rm H}}.
\end{equation}
%
This results in a local timestep
%
\begin{equation}
  \label{eqn:vifront4}
  dt_P = \frac{I}{|dI/dt|} = C_{\rm RT,cfl} \frac{n_{\rm H}}{\tau \dot{n}_{\rm H}},
\end{equation}
%
where $C_{\rm RT,cfl}$ is a safety factor that restricts the change in
intensity to its inverse.  In practice, we have found that a ceiling
of 3 can be placed on the optical depth, so optically thick cells do
not create a very small timestep.  We still find excellent agreement
with analytical solutions with this approximation.  We show the
accuracy using this timestep method in \S\ref{sec:dt_dep}.
